Asymmetry in Cricket: The Effect of Winning the Toss on Winning the Match\thanks{Data and scripts needed to replicate the results are available at: https://github.com/outside-edge/asymmetry/}

### 2023-09-04 ms/ms.tex
replicate (broad sense) + revise ms.

\documentclass[11pt,  letterpaper]{article}

\usepackage{mathpazo}
\usepackage{fontspec}
\setmainfont{TeX Gyre Pagella}

\input{boilerplate}

\title{Asymmetries in Cricket: The Effects of Winning the Toss on Winning the Match\thanks{This paper incorporates some of the text and results from Sood \& Willis's (2019) working paper: ``Fairly Random: The Impact of Winning the Toss on the Probability of Winning.'' The authors thank Matt Blackwell, Andrew Gelman, Donald Green, Jens Hainmueller, and Daniel Stone for their comments.}}

\author{
Apoorva Lal\thanks{Dept.~of Political Science, Stanford University;  \textsf{apoorval@stanford.edu}.} \;\;
Derek Willis\thanks{News applications developer at ProPublica;  \textsf{dwillis@gmail.com}.}  \;\;
Gaurav Sood\thanks{Independent researcher;  \textsf{gsood07@gmail.com}.} \;\;
Avidit Acharya\thanks{Dept.~of Political Science and Graduate School of Business, Stanford University; \textsf{avidit@stanford.edu}.}
}
\date{\today}

%\usepackage[
%  backend=biber,
%  style=authoryear,
%  citestyle=authoryear,
%  ]{biblatex}
%\addbibresource{biblio.bib}
%\renewbibmacro{in:}{}

@@

\maketitle

@@

In cricket, which team bats first is determined by a toss. Does the team that wins the toss have an advantage over the other team? We find that, on average, the effect of winning the toss on winning the match is essentially nil in international limited-overs daytime matches but substantial (3.3 percentage points) in day/night matches. In Test cricket, the advantage is even larger (4.9 percentage points). The advantage has exploded in the last decade (to nearly 15 percentage points) after a steady decline. Surprisingly, the effect of the toss does not appear to vary between competitive and non-competitive pairings of teams. And as for the decision to bat or field first, we find the win-loss differential is significantly larger for teams that elect to field; the advantage is negligible among those that elect to bat. Lastly, and perhaps most surprisingly, in rain-delayed matches in which the chasing team's total is adjusted by the Duckworth-Lewis formula, we find that teams that elected to bat first did just as well as those that elected to field first.

@@

%\textbf{JEL codes:} C72
\end{abstract}

% #### ##    ## ######## ########   #######
%  ##  ###   ##    ##    ##     ## ##     ##
%  ##  ####  ##    ##    ##     ## ##     ##
%  ##  ## ## ##    ##    ########  ##     ##
%  ##  ##  ####    ##    ##   ##   ##     ##
%  ##  ##   ###    ##    ##    ##  ##     ##
% #### ##    ##    ##    ##     ##  #######

@@

In competitive sports, every little thing matters. Nevertheless, many sports leave some key levers out of teams' control and in the hands of fate. In cricket, one such lever---the toss---has been the subject of much attention. In many matches, commentators and cricket enthusiasts claim that there is a definite advantage to either bowling or batting first.\footnote{See, for example, the \href{https://www.quora.com/How-much-does-a-toss-matter-to-a-game-of-cricket}{Quora page on ``How much does a toss matter to a game of cricket?''}} The advantage is also often pointed to by the captains in the pre-toss interview and by the captain of the losing team in the post-match interview. To wrest the advantage, the captain merely needs to pick the side of the coin that will be left facing the sky after the toss and choose whichever one---batting first or batting second---gives the greater advantage.

In this paper, we measure how well teams have been able to capitalize on any advantage conferred by winning the toss. To answer the question, we collected data on over 35,000 cricket matches across four formats (First Class, ODIs, T20s, and Tests) played over the course of more than one hundred fifty years.

We find that winning the toss increases the chances of winning the match by 1.8 percentage points: toss winners prevail 39.2\% of the time and lose 37.4\% of the time; the rest of the matches are drawn or tied. The effect of winning the toss is the largest in Tests, where it is 4.9 percentage points and the smallest in ODIs and T20s, where it is essentially zero on average. That said, in day/night ODIs and T20s, the effect of winning the toss is 3.3 percentage points; in day-time matches, the effect is a precise zero.

The effect of the toss in ODIs has remained relatively stable since the 1980's when the one-day format first became popular. In Tests, however, the importance of the toss has varied considerably. The effect of winning the toss in Test cricket was a little over five percentage points before 1990. It then dropped to almost 0 from 1990 to 2010. In the last decade, however, it has exploded, averaging 14.5 percentage points.

@@

With this, we turn to the batting first decision. When pitch and weather conditions are hard to read, and it is not obvious that fielding first has a clear advantage, many cricket commentators suggest that batting first is the appropriate choice: get runs on the board, and put the opposing team under psychological pressure by having them chase the target, worrying about ``required run rates,'' etc. We see that the effect of winning the toss on batting first is 13 percentage points. If one team bats, the other team must field, so this translates to saying that among toss winners, 56.5\% of the teams elect to bat, and the remaining 43.5\% choose to field. Thus, compared to the case where teams are randomly assigned to bat, i.e., the case in which the toss has no effect on the batting decision, toss-winning teams bat an excess of 6.5\% of matches.

However, we find that of the toss winners that elected to bat, only 38.1\% won the match compared to 37.9\% that lost--- a small difference of 0.2 percentage points. On the other hand, of the toss winners that elected to field, 40.8\% won the match compared to 36.9\% that lost--- a larger difference of 3.9 percentage points. This means that 94\% of the 1.8 percentage point effect of winning the toss can be attributed to instances in which the toss winning team elected to field, and only 6\% of the effect can be attributed to the toss winning team electing to bat. This suggests that teams are better able to capitalize on the toss by electing to field when it is advantageous to do so.

We also look at how the toss advantage breaks down between teams that elected to bat versus those that elected to field within the various subgroups mentioned above and others. Perhaps the most striking result from this analysis for many cricket followers is that there does not appear to be the case that toss winners that elect to bat first do better than those that elect to field first in rain delayed matches where the chasing total is adjusted according to the \citet{duckworth1998} formula.

Lastly, a caveat about our findings. Our empirical approach estimates the extent to which teams have been able to capitalize on the advantage granted by winning the toss. The toss may matter a lot in principle, but teams may decide poorly. So, when we refer to ``toss effects'' in this paper, we do not refer to whether there is an advantage to winning the toss in principle, whether any such advantage is exploited in practice.

Our results are pertinent to debates about how cricket should be governed. One argument is that since the toss is random, no team is advantaged by the format prior to the toss; and in fact the luck of the toss along with the strategic choice to bat first or second are integral parts of the game--- not very different from the luck of weather affecting the teams differently.

An alternative argument, however, is that the toss is problematic because it advantages one team over the other even before the first ball is bowled; and care should be taken to mitigate this asymmetry--- for example, by making cricket pitches more durable so that they don't deteriorate so much so that batting on the final day of a test match is miserable. Otherwise, the toss-winning team could exploit this advantage by batting first, reducing the chance that it has to endure this misery. Moreover, the fact that no side is advantaged \emph{ex-ante} if the toss is random is not enough to justify not correcting any asymmetry created by the toss. In fact, the rules of cricket already try to correct the asymmetry between teams that bat first and those that bat second in one particular instance: In rain-delayed one-day matches, the Duckworth-Lewis formula is used to set a new target for the chasing team.\footnote{The argument that no side was advantaged \emph{ex-ante} could have been used to justify not doing any correction in such rain-delayed matches, and simply declaring that the team that batted first won the match because the chasing team was unable to reach its target. But the designers of one-day cricket were clearly unwilling to allow the bad luck of weather to have this much influence on the game. If addressing the misfortunes of weather are important, it may also be important to address the misfortune of losing the toss.}

For those who take this latter viewpoint, a straightforward policy prescription that comes out of our analysis is that, at least for ODIs, day-night matches should be avoided. In general, when new formats are introduced (such as the recent introduction of T20s) care should be taken to ensure the toss has little effect. The designers of cricket should always consider ways of leveling the playing field for the competing teams starting from the point at which the game really begins--- the moment the first ball is bowled.

We do not address these optimal design questions in this paper, but our results are relevant to these issues depending on what the designers of cricket seek to optimize. However, at the end of the paper, we propose (and discuss the possible consequences of) a subtle change to the rules governing the toss: Rather than tossing a coin to confer the choice to bat first or second upon the captain of the toss-winning team, umpires may consider simply tossing a coin to determine which team must bat first.

% ########     ###    ########    ###
% ##     ##   ## ##      ##      ## ##
% ##     ##  ##   ##     ##     ##   ##
% ##     ## ##     ##    ##    ##     ##
% ##     ## #########    ##    #########
% ##     ## ##     ##    ##    ##     ##
% ########  ##     ##    ##    ##     ##

@@

\subsection{Creating the sample}

We obtained data on cricket matches from the ESPNCricinfo website, which covers international and domestic matches, and maintains statistics on a wide variety of matches across formats.\footnote{Information on the matches was downloaded in June 2020 using a purpose-built library available at \href{https://github.com/dwillis/python-espncricinfo}{https://github.com/dwillis/python-espncricinfo}. The data is available at \href{https://github.com/dwillis/cricket-stats}{https://github.com/dwillis/cricket-stats}.}

We focus on four formats: T20, ODI, Test, and First Class for both men's and women's teams.\footnote{We exclude List A matches that are not ODIs or T20s. These matches are considered a less professional category (with worse data recording) than the formats that we include.} The data are a near census of the relevant population, only excluding scheduled matches that were abandoned without the toss being conducted, and a small number of limited-overs matches in which the minimum number of overs that must be bowled to establish a result were not met.\footnote{In ODIs, for example, each side must bat at least 20 overs for a result to be declared. We have no reason to suspect that the small matches that were excluded because they did not meet this threshold could be biased enough to affect our conclusions. We drop 426 matches recorded as ``no result,'' 3 matches that are recorded as ``abandoned'' or ``canceled,'' and another three that don't record who batted first. The "no result" matches constitute the largest exclusion we make. They appear to be comprised of matches canceled for idiosyncratic reasons, ranging from the \href{https://www.espncricinfo.com/series/shell-tri-series-1991-92-61203/australia-women-vs-england-women-final-66975/full-scorecard}{a team being unable to bat because of the weather} to cancellations because of \href{https://www.espncricinfo.com/series/australia-tour-of-pakistan-1982-83-61392/pakistan-vs-australia-3rd-odi-64196/full-scorecard}{unwanted audience participation}. In all, we drop approximately 0.07\% of matches.}

Our final sample comprises 35,357 cricket matches across four formats played over the last 150 years. We report the distribution of matches over time and by type in Figure \ref{fig:timeseries_decomp}. As the figure shows, the number of matches has exploded over the last two decades, largely due to the introduction of the enormously popular Twenty-20 (T20) format of the game both in international and domestic tournaments. T20 matches constitute over a quarter of all matches played in the last decade.

% subsubsection data_collection (end)

\subsection{Is the toss random?}

Since our analysis rests on the assumption that the coin toss is random, we start by examining this assumption.\footnote{Prior work by \citet{shafqat2015analysis} and \citet{dawson2009bat} conditions on post-treatment variables, whereas we use only the randomization of the toss to identify toss effects.} Is there any evidence that the coin toss is rigged---for example, does the home side enjoy the rub of the green more often? In order to study this, we compare the rates at which home teams, teams that won the toss in their last match, and underdogs (defined as the lower-ranked team whenever rankings are available for both teams) win the toss relative to their opponents.

To examine whether the toss is rigged, we compute the odds of seeing as many successes in the same number of draws of a sequence of independent Bernoulli random variables with parameter 0.5. For example, home teams have won 50.81\% of tosses in the 6,377 matches for which we have data. The likelihood of tossing a fair coin 6,377 times and getting at least this percentage of heads is 20.1\%. This chance appears low, but not eyebrow-raisingly so. We find that toss wins are even more balanced for teams that won and lost the last toss, and for teams that are underdogs and favorites. Table \ref{table: balance} reports these balance results.

That said, we do see one potential source of concern. We find that teams that share their nationality with at least two of the three umpires of the match win 51.3\% tosses in the 4,044 international matches for which we have this data. The likelihood of tossing a fair coin 4,044 times and getting at least this share of heads is only 10\%. This number is certainly low enough to raise some eyebrows. That said, the practice of allowing home-team umpires in international matches ended in 1992, and so these issues are less relevant in recent decades which comprise the bulk of our sample. In addition, a rigged toss in favor of the home team would bias estimates of the advantage of winning the coin toss to the extent that it is correlated with ability. If stronger teams win more tosses, estimates of the advantage of the coin toss would be inflated upwards. And vice versa, if otherwise. Given home teams win more matches and are seen as 'stronger' at home, it is possible that the results are slightly biased upwards. We investigate this point later in the paper, looking at effects post 1992. For now, we proceed as if the toss was random.

% ########  #######   ######   ######
%    ##    ##     ## ##    ## ##    ##
%    ##    ##     ## ##       ##
%    ##    ##     ##  ######   ######
%    ##    ##     ##       ##       ##
%    ##    ##     ## ##    ## ##    ##
%    ##     #######   ######   ######

\section{Do teams exploit the advantage of winning the toss?}

To estimate the advantage of winning the toss, we run regressions on a team $\times$ match level dataset. Let $i$ index teams, and $j$ index matches. $Y_{ij}$ is a dummy for team $i$ winning match $j$, and takes a value of $0.5$ for draws. $Z_{ij}$ is a dummy for team $i$ winning the toss in match $j$. For the main specifications of interest, we estimate regression equations of the form

@@

We report results in Table \ref{table:reduced_form}. We progressively introduce team, match-type, and time period fixed effects in columns 2, 3, and 4, and find that results are remarkably stable. The first column gives the baseline estimate of the toss effect. The 1.8 percentage points difference reflects the fact that toss winners win 39.2\% of the time and lose 37.4\% of the time; the remaining matches are drawn. The estimated effect remains at 1.7 percentage points when we progressively add various fixed effects.

We then proceed to examine the effect of the toss within various subgroups. We report the estimates graphically in Figures \ref{fig:rf_het_TE} - 4 with point estimates and sample sizes as annotations. We summarize the results below:

\textbf{Format.} Toss effects are most pronounced in Test matches followed by first class matches, and are statistically indistinguishable from zero in ODIs and T20s. We estimate the effect of the toss on winning a Test match to be 4.7 percentage points. The effect of the toss in first class matches is 2.3 percentage points. These results support the conventional wisdom among lay cricket followers that the toss grants the greatest advantage in multi-day affairs. Pitches invariably deteriorate over the course of these matches, and the fourth innings in a Test match is often the most challenging time to bat. The pitch deteriorates far less over the course of a day, or in the case of T20s, a few hours. Our null results for limited over matches also extend \citet{de1998winning}, who, based on analysis of data from 427 international one-day matches conclude that ``winning the toss at the outset of a match provides no competitive advantage.''

\textbf{International vs.~Domestic matches.} Toss effects are virtually the same for international and domestic matches (point estimates are 1.7 percentage points in the former, and 1.9 percentage points in the latter).

\textbf{Effects over time.} Toss effects appear to have declined over time from a high of 9.9 percentage points in the second half of the 19th century to 1.4 percentage points in the last decade. This trend may be a result of a changing composition of matches so we also looked at these trends within formats. T20s have not been played long enough for us to draw any conclusions about how toss effects have changed over time. But within the other formats, we find that toss effects follow a similar pattern, except that in Test cricket there is a peculiar uptick in the last decade: the toss effect jumps from 0 in the decade before to 14.5 percentage points now. We investigate potential sources of the ecological decline below but there are some factors that could be important drivers but for which we do not have the data. For example, cricketers may have become fitter, and able to play for longer periods without tiring as much as they once did, reducing the impact of the order in which they ball or bat. Pitches may have become more resilient to wear than they once were. Teams may have become better informed about how to play more effectively whether they bat first or second, etc.

\textbf{Relative strength.} We expect the toss to be less important in matches where one team is clearly stronger than the other. We expect the stronger team to win, whether it bats first or second. To test this hypothesis, we use team ranking data published by cricket's official governing body, the International Cricket Council (ICC). The median match is played between teams where the ICC rating difference is 17 points. We use that as the approximate splitting point, splitting the matches into teams where the gap in the ICC ratings of the two teams is less than 15 points and where it is more. In this case, we believe the standard errors are too large for us to arrive at any substantively meaningful conclusions. Statistically, we believe effects larger than 8\% are unlikely.

\textbf{Continents.} Toss effects vary only somewhat across continents, hovering around 2.5 percentage points for matches played in Asia, Oceania, and the Americas (though the estimates for the Americas are noisier in part due to the smaller number of matches played there). In Europe (effectively England) the toss effect is only 1.1 percentage points (though we cannot rule out a null effect), and in Africa it is essentially zero. Overall, despite there being much talk about the different kinds of wickets across the world, with wickets in certain regions being drier, or stickier, or subject to more wear, these differences do not seem to translate into substantial toss effect differences at least when we look at coarse geographic averages.

\textbf{Day/Night matches.} It is often claimed that the toss is more crucial in day-and-night matches. We find some evidence for this. In the limited overs formats where day/night matches take place (ODIs and T20s), the advantage of winning the toss in a day/night match is 3.3 percentage points, whereas the

advantage of winning the toss in a limited overs match played during the day is zero.

\textbf{Seasonality.} Weather is also thought to have a large impact on the playing conditions in cricket. Cool overcast weather, for example, is thought to aid swing bowling, especially on English pitches. More generally, the advantage of winning the toss likely varies by weather. Although we do not have data on the weather, we can proxy it with seasons. In particular, students of the game suspect that the advantage of winning the toss in the early English season is especially significant. There is some evidence of a seasonal pattern for matches played in England, with the advantage of winning the toss growing over the summer, though the trend is weak and broken in the month of July.

\textbf{Rain delays.} Finally, we look at the differential effects of winning the toss in ODI matches that were rain-delayed, causing the number of overs bowled in the second innings to be adjusted down. We examine the subset of matches where rain-delays result in revised targets computed using the \citet{duckworth1998} method. This method has been the subject of much controversy, with many cricket fans suggesting that it is rigged against the chasing team. To our surprise, we find no evidence that toss winners are able to exploit the possibility of second-inning rain delays to their advantage. If anything, we see that the toss-winner has a \emph{disadvantage} of 2.4 percentage points in matches subject to the Duckworth-Lewis adjustment, though this point estimate is statistically indistinguishable from zero.\footnote{Toss-winners won 160 out of 333 ODIs with the Duckworth-Lewis adjustment (i.e.~48\%) and 3085 out of 6187 ODIs without the Duckworth-Lewis adjustment (49.9\%).} This, however, does not mean that the Duckworth-Lewis method is not rigged against the chasing team; that issue will be discussed further in the next section.

% subsubsection heterogeneous_treatment_effects (end)

% ########     ###    ########
% ##     ##   ## ##      ##
% ##     ##  ##   ##     ##
% ########  ##     ##    ##
% ##     ## #########    ##
% ##     ## ##     ##    ##
% ########  ##     ##    ##

\section{Batting versus Fielding First}

>>>>

\documentclass[12pt, letterpaper]{article}
\usepackage[titletoc,title]{appendix}
\usepackage{color}
\usepackage{booktabs}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{dark-red}{rgb}{0.75,0.10,0.10}
\definecolor{bluish}{rgb}{0.05,0.05,0.85}
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage[linkcolor=blue,
			colorlinks=true,
			urlcolor=blue,
			pdfstartview={XYZ null null 1.00},
			pdfpagemode=UseNone,
			citecolor={bluish},
			pdftitle={social proof}]{hyperref}

%\newcites{SI}{SI References}
\usepackage{natbib}

\usepackage{float}
\usepackage{placeins}

\usepackage{geometry}  % see geometry.pdf on how to lay out the page. There's lots.
\geometry{letterpaper} % This is 8.5x11 paper. Options are a4paper or a5paper or other...
\usepackage{graphicx}  % Handles inclusion of major graphics formats and allows use of
\usepackage{amsfonts,amssymb,amsbsy}
\usepackage{amsxtra}
\usepackage{verbatim}
%\setcitestyle{round,semicolon,aysep={},yysep={;}}
\usepackage{setspace} % Permits line spacing control. Options are:
%\doublespacing
%\onehalfspace
%\usepackage{sectsty}    % Permits control of section header styles
\usepackage{pdflscape}
\usepackage{fancyhdr}   % Permits header customization. See header section below.
\usepackage{url}        % Correctly formats URLs with the \url{} tag
\usepackage{fullpage}   %1-inch margins
\usepackage{multirow}
\usepackage{verbatim}
\usepackage{rotating}
\setlength{\parindent}{3em}

%\usepackage[T1]{fontenc}
%\usepackage[bitstream-charter]{mathdesign}

\usepackage{chngcntr}
\usepackage{longtable}
\usepackage{adjustbox}
\usepackage{dcolumn}

\usepackage[nameinlink, capitalize, noabbrev]{cleveref}

\def\citeapos#1{\citeauthor{#1}'s (\citeyear{#1})}

\makeatother

\usepackage{footmisc}
\setlength{\footnotesep}{\baselineskip}
\makeatother
\renewcommand{\footnotelayout}{\normalsize \doublespacing}

% Colors
\usepackage{color}

\newcommand{\bch}{\color{blue}\em  }   % begin change
\newcommand{\ying} {\color{orange}\em  }   % begin change
\newcommand{\bgcd} {\color{purple}\em }
\newcommand{\ech}{\color{black}\rm  }    % end change

% Caption
\usepackage{caption}
%\captionsetup[subtable]{font=small,skip=0pt}
\usepackage{subcaption}

% tt font issues
% \renewcommand*{\ttdefault}{qcr}
\renewcommand{\ttdefault}{pcr}

\usepackage{tocloft}

\usepackage{lscape}
\renewcommand{\textfraction}{0}
\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}
\renewcommand{\floatpagefraction}{0.40}
\setcounter{totalnumber}{5}
\makeatletter
\providecommand\phantomcaption{\caption@refstepcounter\@captype}
\makeatother

\usepackage{footmisc}
\renewcommand{\footnotelayout}{\setstretch{2}}

\title{Asymmetry in Cricket: The Effect of Winning the Toss on Winning the Match\thanks{Data and scripts needed to replicate the results are available at: https://github.com/outside-edge/asymmetry/}}

\author{
Apoorva Lal\thanks{Data Scientist, Netflix;  \href{mailto:lal.apoorva@gmail.com}{lal.apoorva@gmail.com }.} \;\;
Derek Willis\thanks{Lecturer in Data and Computational Journalism, University of Maryland;  \href{mailto:dwillis@gmail.com}{dwillis@gmail.com}.}  \;\;
Gaurav Sood\thanks{Independent researcher;  \href{mailto:gsood07@gmail.com}{gsood07@gmail.com}; Corresponding author. Phone: 650 269 7946. Address: 156th Place NE, Redmond, WA, 98052.}\;\;
Avidit Acharya\thanks{Department of Political Science and Graduate School of Business, Stanford University; \href{mailto:avidit@stanford.edu}{avidit@stanford.edu}.}
}
\date{\today}

@@

\maketitle
\doublespacing

@@

In cricket, which team bats first is determined by a toss. Does the team that wins the toss do better? We find that, on average, the effect of winning the toss on winning the match is essentially nil in international limited-overs daytime matches but substantial (3.3 percentage points) in day/night matches. In Test cricket, the advantage is yet larger (4.9 percentage points). Surprisingly, the effect of winning the toss does not vary substantially across matches between evenly matched and mismatched teams. Lastly, and perhaps most surprisingly, in rain-delayed matches in which the chasing team's total is adjusted by the Duckworth-Lewis formula, we find that the effect of winning the toss is statistically indistinguishable from zero.

@@

\end{abstract}

@@

In competitive sports, teams fight for every little advantage. Many sports, however, leave some important things to chance. In cricket, a coin toss is used to decide which team bats first. The choice is widely seen as consequential. In many matches, commentators and cricket enthusiasts claim that there is a definite advantage to either bowling or batting first.\footnote{See, for example, the \href{https://www.quora.com/How-much-does-a-toss-matter-to-a-game-of-cricket}{Quora page on ``How much does a toss matter to a game of cricket?''}.} The advantage is also often pointed to by the captains in the pre-toss interview and by the captain of the losing team in the post-match interview.

In this paper, we measure the advantage conferred by winning the toss. To answer the question, we collected data on over 35,000 cricket matches played over 150 years---the largest ever dataset assembled for the question, and more than 50 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}. In analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into. To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on post coin-toss decisions.

We find that winning the toss increases the chances of winning the match by 1.8 percentage points: toss winners prevail 39.2\% of the time and lose 37.4\% of the time; the rest of the matches are drawn or tied. The effect of winning the toss is the largest in Tests, where it is 4.9 percentage points, and the smallest in ODIs and T20s, where it is essentially zero on average. That said, in day/night ODIs and T20s, the effect of winning the toss is 3.3 percentage points.

@@

With this, we turn to what teams choose after winning the toss. When pitch and weather conditions are hard to read, many cricket commentators suggest that batting first is the appropriate default choice: get runs on the board and put the opposing team under psychological pressure by having them chase the target. That advice is borne out in the data. Toss winners elect to bat somewhat more often (56.5\% of the time) than they elect to field. However, of the toss winners that elect to bat, only 38.1\% win the match compared to 37.9\% that lose. On the other hand, of the toss winners that opt to field, 40.8\% win the match compared to 36.9\% that lose. This means that 94\% of the 1.8 percentage point effect of winning the toss can be attributed to instances in which the toss-winning team elected to field, and just 6\% of the effect can be attributed to the toss-winning team electing to bat.

Lastly, a caveat about our findings. Our empirical approach estimates the extent to which teams have capitalized on the advantage granted by winning the toss and not the theoretical advantage granted by the toss. The toss may matter a lot in principle, but teams may decide poorly.

The rest of the paper is organized as follows. We start by describing the data. We then analyze if the data are consistent with the toss being random. Next, we present the results. We end with discussing the policy implications.

@@

We collected data on cricket matches from ESPNCricinfo using a purpose-built library \citep{Willis_python-espncricinfo_A_Python_2022} in June 2020. From the data we collected, we discarded List A matches that were not ODIs or T20s because they are considered a less professional category (with worse data recording). This leaves us with data on all the men's and women's cricket matches in four formats: T20, ODI, Test, and First Class. From this census of the relevant population, we discard matches that were abandoned without the toss being conducted, and a small number of limited-overs (ODI and T20) matches in which the minimum number of overs that must be bowled to establish a result were not bowled. (In ODIs, for example, each side must bat at least 20 overs for a result to be declared.) In all, we dropped 426 matches with ``no result,'' 3 matches that were ``abandoned'' or ``canceled,'' and another three where we have no record for who batted first. The data also suggest little reason to be concerned about biased attrition. The ``no result'' matches, which constitute the overwhelming majority of matches we discard, comprise of matches that were canceled for idiosyncratic reasons, ranging from the \href{https://www.espncricinfo.com/series/shell-tri-series-1991-92-61203/australia-women-vs-england-women-final-66975/full-scorecard}{a team being unable to bat because of the weather} to cancellations because of \href{https://www.espncricinfo.com/series/australia-tour-of-pakistan-1982-83-61392/pakistan-vs-australia-3rd-odi-64196/full-scorecard}{unwanted audience participation}.

Our final sample comprises 35,357 cricket matches across four formats played over 150 years. We report the distribution of matches over time and by type in Figure \ref{fig:timeseries_decomp}. As the figure shows, the number of matches has exploded over the last two decades. This is largely due to the introduction of the enormously popular T20 format of the game; T20 matches constitute over a quarter of all matches played in the last decade.

\section{Is The Toss Random?}

Since our analysis rests on the assumption that the coin toss is random, we start by examining this assumption. Is there any evidence that the coin toss is rigged? For example, does the home side enjoy the rub of the green more often? To study this, we compare the rates at which home teams, teams that won the toss in their last match, and underdogs (defined as the lower-ranked team whenever rankings are available for both teams) win the toss relative to their opponents. We compute the odds of seeing as many successes in the same number of Bernoulli trials with a probability of 0.5. For example, home teams have won 50.8\% of tosses in the 6,377 matches for which we have data. The likelihood of tossing a fair coin 6,377 times and getting as large a percentage of heads is 20.1\%. We find that toss wins are even more balanced for teams that won (lost) the last toss, and for teams that are underdogs and favorites. Table \ref{table: balance} reports these balance results.

That said, we do see one potential source of concern. We find that teams that share their nationality with at least two of the three umpires of the match win 51.3\% of the tosses in the 4,044 international matches for which we have this data. The likelihood of tossing a fair coin 4,044 times and getting this large a share of heads is only 10\%. This number is certainly low enough to raise some eyebrows. That said, the practice of allowing home-team umpires in international matches ended in 1992, and so these issues are less relevant in recent decades which comprise the bulk of our sample. In addition, a rigged toss in favor of the home team would bias estimates of the advantage of winning the coin toss to the extent that it is correlated with ability. If stronger teams win more tosses, estimates of the advantage of the coin toss would be inflated upwards. And vice versa, if otherwise. Given that home teams win more matches and are seen as stronger, it is possible that the results are biased upwards. For now, we proceed as if the toss was random.

\section{The Effect of Winning the Toss on the Probability of Winning}

To estimate the advantage of winning the toss, we run regressions on a team $\times$ match-level dataset. Let $i$ index teams, and $j$ index matches. $Y_{ij}$ is a dummy for team $i$ winning match $j$, and takes a value of $0.5$ for draws. $Z_{ij}$ is a dummy for team $i$ winning the toss in match $j$. For the main specifications of interest, we estimate regression equations of the form

@@

We report results in Table \ref{table:reduced_form}. The first column gives the baseline estimate of the toss effect. We progressively add team, match-type, and time period fixed effects in columns 2, 3, and 4, and find that the results are remarkably stable. The baseline estimate of 1.8 percentage points reflects the fact that toss winners win 39.2\% of the time and lose 37.4\% of the time; the remaining matches are drawn. The estimated effect shuffles to 1.7 percentage points when we add various fixed effects.

We next examine the effect of the toss across various subgroups. We report the estimates graphically in Figures \ref{fig:rf_het_TE} - \ref{fig:rf_het_TE3} with point estimates and sample sizes as annotations. We summarize the results below:
\begin{itemize}
\item \textbf{Format.} Toss effects are most pronounced in Test matches followed by first-class matches, and are statistically indistinguishable from zero in ODIs and T20s. The effect of the toss on winning a Test match is 4.7 percentage points and 2.3 percentage points in first-class matches. These results support the conventional wisdom among lay cricket followers that the toss grants the greatest advantage in multi-day affairs. Pitches invariably deteriorate over the course of these matches, and the fourth innings of a Test match is often the most challenging time to bat. The pitch deteriorates far less over the course of a day or in the case of T20s, a few hours. Our null results for limited over matches also extend \citet{de1998winning}, who, based on analysis of data from 427 international one-day matches conclude that ``winning the toss at the outset of a match provides no competitive advantage.''

\item \textbf{International vs.~Domestic matches.} Toss effects are virtually the same for international and domestic matches (point estimates are 1.7 percentage points in the former and 1.9 percentage points in the latter).

\item \textbf{Effects over time.} Toss effects appear to have declined over time from a high of 9.9 percentage points in the second half of the 19th century to 1.4 percentage points in the last decade. This trend may be a result of a changing composition of matches so we also looked at trends within formats. T20s have not been played long enough for us to draw any conclusions about how toss effects have changed over time. But in the other formats, except for Test cricket, where there has been an uptick in the last decade, we find a similar pattern as above. The potential reasons for the decline are numerous. For example, cricketers may have become fitter, and more able to play for long periods without tiring as much as they once did, reducing the impact of the order in which they batted. Pitches may have become more resilient to wear than they once were. Teams may have become better informed about how to play more effectively whether they bat first or second, etc.

\item \textbf{Relative strength.} We expect the toss to be less important in matches where one team is clearly stronger than the other. We expect the stronger team to win, whether it bats first or second. To test this hypothesis, we use team ranking data published by cricket's official governing body, the International Cricket Council (ICC). The median difference between teams is 17 points. We use that as an approximate splitting point, splitting the matches into teams where the gap in the ICC ratings of the two teams is less than 15 points and where it is more. In this case, we believe the standard errors are too large for us to draw substantively meaningful conclusions.

\item \textbf{Continents.} Toss effects vary somewhat across continents, hovering around 2.5 percentage points for matches played in Asia, Oceania, and the Americas (though the estimates for the Americas are noisier in part due to the smaller number of matches played there). In Europe (effectively England) the toss effect is only 1.1 percentage points (though we cannot rule out a null effect), and in Africa, it is essentially zero.

\item \textbf{Day/Night matches.} It is often claimed that the toss is more crucial in day-and-night matches. Part of the claim rests on the theory that batting is easier when there is dew, which generally comes in the evening. Another popular theory is that the ball used for day/night limited overs cricket is harder to see at night. We do have the data to adjudicate between the theories but have data that shows that the toss is more consequential in day/night limited over games. In the limited overs formats where day/night matches take place (ODIs and T20s), the advantage of winning the toss is 3.3 percentage points, compared to 0 for limited overs matches during the day.

\item \textbf{Seasonality.} Weather is thought to have a large impact on the playing conditions in cricket. Cool overcast weather, for example, is thought to aid swing bowling, especially on English pitches. More generally, the advantage of winning the toss likely varies by weather. Although we do not have data on the weather, we can proxy it with seasons. In particular, students of the game suspect that the advantage of winning the toss in the early English season is especially significant. There is some evidence of a seasonal pattern for matches played in England, with the advantage of winning the toss growing over the summer, though the trend is weak and broken in the month of July.

\item \textbf{Rain delays.} Finally, we look at the differential effects of winning the toss in ODI matches that were rain-delayed, causing the number of overs bowled in the second innings to be adjusted down. We examine the subset of matches where rain delays result in revised targets computed using the \citet{duckworth1998} method. This method has been the subject of much controversy, with many cricket fans suggesting that it is rigged against the chasing team. To our surprise, we find no evidence that toss winners are able to exploit the prediction of rain delays to their advantage. If anything, we see that the toss-winners have a 2.4 percentage points \emph{disadvantage} in matches subject to the Duckworth-Lewis adjustment, though this point estimate is statistically indistinguishable from zero.\footnote{Toss-winners won 160 out of 333 ODIs with the Duckworth-Lewis adjustment (i.e.~48\%) and 3,085 out of 6,187 ODIs without the Duckworth-Lewis adjustment (49.9\%).}

\end{itemize}

\section{Batting Vs. Fielding First}

56.5\% of teams that won the toss elected to bat. This is not surprising given the conventional wisdom is that when you are not sure about whether it is advantageous to bat or field first, bat first. The first column of Table \ref{table:iv_table} shows that 49.2\% of teams that batted first won the match, while 50.8\% of teams that fielded first won. The second column shows that among teams that won the toss and batted first the ``win-loss differential'' is small: these teams won only 0.2 percentage points more frequently than they lost, winning 38.1\% of matches and losing 37.9\% of matches, as reported in the introduction. This win-loss differential is smaller than it is for teams that won the toss and elected to field. The third column of the table shows that these teams won 3.9 percentage points more frequently than they lost: they won 40.8\% of their matches and lost 36.9\% of them. Combining this information with the fact that 56.5\% of teams that won the toss elected to bat (and only 43.5\% elected to field, as implied by the fourth column of the table), we learn that 94\% ($= .435 \times .039/.018$) of the 1.8 percentage point advantage to winning the toss that we reported in the previous section can be attributed to the success of toss-winning teams that elected to field. To further understand what explains the preference to bat or bowl first, we see how winning the toss affects the choice in various subsets:

\begin{itemize}
\item \textbf{Format.} Teams prefer batting first in longer match formats. In Test matches 72.8\% of teams that win the toss bat first, whereas in T20s and ODIs this share is only about 52.9\%. In first-class matches it is 56.9\%.
